---
title: "04 Import and transform data"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Import and transform data with the tidyverse.
---

```{r setup, include=FALSE}
library(learnr)
library(intRo)
data("pyramids")
library(tidyverse)
fruits <- read_csv(system.file("extdata", "fruits.csv", package = "intRo"))
knitr::opts_chunk$set(echo = TRUE)
```


## Import data

So far we've been using ready-made data.
However, it's most common for data to be imported from files into R.

Importing data in R is quite straightforward.
Most of the data we use is in the form of tables: you have rows, columns, and values in each cell. Like an Excel spreadsheet.
You will learn how to import tabular data now, but we will see how to import other kinds of data tomorrow.

### Tabular data

The simplest and most common way of storing tabular data is with *comma separated values* files. These files are simple text files, where each row has a set of values separated by commas `,`.

The commas separate the columns of the table. Look at the following example:

```
fruit,quantity,provenance
orange,2,Spain
banana,4,Brazil
apple,1,UK
blueberry,15,UK
```

The first row is called the *header*. The header specifies the names of the columns in the file: here the columns are two and they are `fruit`, `quantity` and `provenance`. The column names are separated by a comma.

The other rows have the values for each column, separated by commas. You can think of each row after the header as one observation.

Comma separated files generally have the `.csv` extension.

To import data from a `.csv` file into R, you can use the `read_csv()` function, from the tidyverse package [readr](https://readr.tidyverse.org). 

First, attach the tidyverse (which includes readr).

```{r tidyverse, exercise=TRUE}
library(tidyverse)
```

`read_csv()` only needs you to specify where the file is, in other words the full path to the file.

Add `read_csv()` below and include between the parenthesis the following path: `"https://bit.ly/3HmuVR5"`.
It's a web link! Yes, you can even read data from the internet directly.

```{r fruits, exercise=TRUE}

```

```{r fruits-hint, echo=FALSE, eval=FALSE}
fruits <- read_csv("")
```


### Inspect your data

After importing the data, you can quickly look at them by simply typing the name of the data frame in the console and pressing enter.

Here you can see how the output looks like.

```{r fruits-insp}
fruits
```



## Filter

**Data transformation** is a fundamental aspect of data analysis.

After you have imported data into R, you might want to filter rows, create new columns, or select some of the existing columns.

We will work with the `pyramids` data frame.
Here's what it looks like.

```{r pyr}
pyramids
```

Now inspect the data.

```{r pyr-inspect, exercise=TRUE}

```


### Filter data

You can filter a data frame with `filter()`, from the dplyr package.

`filter()` needs two things:

- The data frame you want to filter.

- At least one rule to filter the data with.

Let's filter `pyramids` so that it returns only pyramids from the Saqqara site.

Couldn't be easier!

```{r saqq, exercise=TRUE}
filter(
  pyramids,            # The data frame
  Site == "Saqqara"    # The rule
)
```

The code above filters the rows so that only the rows where `Site` equals (`==`) `"Saqqara"` are returned.

```{r n-pyr, echo=FALSE}
question(
  "How many pyramids are there in Saqqara?",
  answer("8"),
  answer("5", correct = TRUE),
  answer("16"),
  answer("2")
)
```

Let's try again now using the pipe `%>%`.
Do you remember how it works?

```{r saqq-pipe, exercise=TRUE}

```

### Logical operators

The double equal sign we used above `==` is called a *logical operator*.
Logical operators are special symbols that allow you to compare objects.

There are the basic logical operators:

- `A == B`: A is *equal* to B.
- `A != B`: A in *not* equal to B.
- `A > B`: A is *greater* than B.
- `A < B`: A is *smaller* than B.

Logical operators return a logical object, either `TRUE` or `FALSE`.

```{r logic}
10 > 4
2 == 3
```

Try the logical operators out! Fill in the following code so that you get TRUE in the first 3 comparisons and FALSE in the last 3.

```{r logic-ex, exercise=TRUE, eval=FALSE}
# These should return `TRUE`
... < 20
"a" == ...
"apple" != ...

# These should return `FALSE`
"zoo" ... "park"
120 ... 43
0.1 ... 0.002
```

### Multiple filtering

You can filter a data frame depending on more than one column.
Simply add more rules, separated by commas.

```{r multi-filt, exercise=TRUE}
pyramids %>%
  filter(
    Site == "Abydos",
    Dynasty > 3
  )
```

```{r aby-3, echo=FALSE}
question(
  "How many pyramids are there in Abydos that were build after the third dynasty?",
  answer("10"),
  answer("0"),
  answer("3", correct = TRUE)
)
```

Now try to find out how many pyramids are there that were built after the tenth dynasty an not in Dahshur.

```{r dahshur-10, exercise=TRUE}
pyramids %>%
  filter(
    ...,
    ...
  )
```


## Mutate

What if you want to create a new column based on existing columns?

You can create new columns with the `mutate()` function from dplyr.

### How to create a new column

`mutate()` needs the name of the data frame you want to mutate, the name of the new column and the code to create it.

```{r square, exercise=TRUE}
pyramids <- pyramids %>%     # Ceci c'est un pipe. ;)
  mutate(
    square = Base1 == Base2
  )
```

Woah! A lot of new stuff going on.
Let's unpack it.

This time we have saved the output of `mutate()` back into `pyramids` using the assignment symbol `<-`, so that we can use the new column in further code.
In other words, we have overwritten `pyramids` with the mutated data frame.

Then we use `mutate()` to create a new column called `square`.

Note that the name of the new column is followed by *one* equal sign `=`, not two.
We are not comparing objects, but creating one.

We are telling `mutate()` to fill the new column with `TRUE` if `Base1` is the same length as `Base2` (the base is squared), with `FALSE` if `Base1` is not the same length as `Base2` (the base is not squared).

To test if `Base1` is as long as `Base2`, we are using the logical operator `==` (two equal signs).
We are comparing two objects here, so we use two equals, not just one.

Check the data now to look at the `square` column.

```{r pyr-square, exercise=TRUE, exercise.setup="square"}
pyramids
```

### Count occurences

We can use the `count()` function to count the number of occurrences for each value of a column.
Let's count how many pyramids have a squared base.

`count()` takes the name of the column you want to count values in.

```{r square-count, exercise=TRUE, exercise.setup="square"}
pyramids %>%
  count(square)
```

Now let's create a new column called `mudbrick`.
The column should have `TRUE` if `Material` is `Mudbrick`, `FALSE` otherwise.

```{r material, exercise=TRUE}
pyramids <- pyramids %>%
  mutate(
    ...
  )
```

```{r mud-count-setup, echo=FALSE}
pyramids <- pyramids %>%
  mutate(
    square = Base1 == Base2,
    mudbrick = Material == "Mudbrick"
  )
```

And now count how many pyramids are build with mudbrick.

```{r mud-count, exercise=TRUE}
pyramids %>%
  ...
```

<div id="mud-count-hint">
**Hint:** Use the `count()` function.
</div>


## Select

What if you only want some columns? Use the `select()` function from dplyr.

`select()` takes a list of unquoted column names (i.e. the bare names, without quotes `"`) separated by commas.

```{r select, exercise=TRUE}
pyramids %>%
  select(Pharaoh, Site)
```

To remove one or more columns, prefix the column name with `-` (minus). All the other columns will be kept.

```{r deselect, exercise=TRUE, exercise.setup="mud-count-setup"}
pyramids %>%
  select(-square, -mudbrick)
```

To select a range of columns you can include only the first and the last of the range, separated by `:`.

```{r select-range, exercise=TRUE}
pyramids %>%
  select(Modern_name:Dynasty)
```


### You did it!

Success! You completed this tutorial.
