---
title: "02 R basics"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = TRUE)
```

## Mathematical operations

Run the following code to see the results.

```{r maths, exercise=TRUE}
14 + 75

76.5 - 13.46

5 * 2598

4209 / 3
```

Now write the code for the following operations.

1. Sum 5 with 8 and multiply by 3 (= 39).

2. Divide 60 by the sum of 5 and 3 (= 7.5).

3. Divide the sum of 76 and 43 multiplied by 5, by the difference between the product of 5 and 3 and the product of 7 and 8 (-14.5122).

```{r paren, exercise=TRUE}
# 1.

# 2.

# 3.

```

## Variables

### Create a variable

Let's try creating some variables.

Assign the value `4` to a variable called `fruit`.
Then subtract 1 from it.

```{r var, exercise=TRUE}
# Create the variable

# Subtract 1 from it

```

### More than one value

We can assign more than a value to a variable, as long as they are all of the same type (like numeric or a string).

```{r combine, echo=FALSE}
question(
  "Which of the following functions can be used to assign multiple values to a variable?",
  answer("`cat()`"),
  answer("`print()`"),
  answer("`c()`", correct = TRUE),
  answer("`t()`")
)
```

Now create a variable `houses` with the values `"Gryffindor", "Slytherin", "Hufflepuff"`.

```{r houses, exercise=TRUE}
houses <- ...
```

### Re-use variables

Oops! We forgot one Hogwarts house.
Let's fix that!

```{r extra, echo=FALSE}
question(
  "Which of the following lines of code let us add an extra value to `houses`?",
  answer('`houses <- c(houses, "Ravenclaw")`', correct = TRUE),
  answer('`houses <- c("Ravenclaw")`'),
  answer('`houses + "Ravenclaw"`')
)
```

Now write the code here and run it to add the missing house!

```{r missing, exercise=TRUE}
houses ...
```

## Functions

```{r function, echo=FALSE}
question(
  "Which of the following is true?",
  answer("A function needs at least one argument."),
  answer("Functions are strings."),
  answer("A function can take one value."),
  answer("Functions can take any number of arguments.", correct = TRUE)
)
```

Run the following code to find out the mean number of fish spotted in a pond over four days.

```{r mean, exercise=TRUE}
fish <- c(1, 6, 2, 10)

sum(fish) / length(fish)

mean(fish)
```

```{r length, echo=FALSE}
question(
  "What is the `lenght()` function for?",
  answer("It calculates the sum of the numbers in the variable."),
  answer("It returns the size of the fish."),
  answer("It returns the number of values in a variable.", correct = TRUE)
)
```

