---
title: "03 Plotting"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(intRo)
data("endangered")
data("slavery_tot")
knitr::opts_chunk$set(echo = FALSE)
```

## The basics

The tidyverse package [ggplot2](https://ggplot2.tidyverse.org) provides ...

```{r tidy-load, exercise=TRUE}
library(tidyverse)
```

These are the minimum constituents of a ggplot.

- The `data`: you have to specify the data frame with the data you want to plot.

- The `mapping`: the mapping tells ggplot how to map data to parts of the plot like the axes or groupings within the data. These parts are called *aesthetics*, or `aes` for short.

```{r}
ggplot(
  data = msleep,
  mapping = aes(
    x = sleep_total,
    y = sleep_rem
  )
)
```


Build your first plot!

```{r first-plot, exercise=TRUE}
ggplot(
  data = ...,
  mapping = aes(
    ...
  )
)
```




### Number of languages by endangerment status

```{r}
library(intRo)
data("endangered")
```


```{r endanger, exercise=TRUE}
endangered %>%
  ggplot(aes(status)) +
  geom_bar()
```


### Number of enslaved people by ship flag

```{r slavery, exercise=TRUE}
library(intRo)
data("slavery")
```


```{r bare, exercise=TRUE}
ggplot(
  data = slavery_tot,
  mapping = aes(country, enslaved)
)
```


```{r bars, exercise=TRUE}
ggplot(
  data = slavery_tot,
  mapping = aes(country, enslaved)
) + geom_bar(stat = "identity")
```


