---
title: "03 Plotting"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(intRo)
data("endangered")
data("slavery_tot")
data("slavery")
data("harry_potter")
data("land_use")
knitr::opts_chunk$set(echo = FALSE)
```

## The basics

### Start here

The tidyverse package [ggplot2](https://ggplot2.tidyverse.org) provides users with a consistent set of functions to create captivating graphics. 
To be able to use the functions in a package, you first need to attach the package.
Here, we can attach the entire tidyverse, which includes ggplot2.

You only have to attach packages once, when you start your R session in RStudio.
They will stay attached until you close RStudio and end the R session.

Attach the tidyverse now.

```{r tidy-load, exercise=TRUE}
library(tidyverse)
```

### A basic plot

These are the minimum constituents of a ggplot.

- The `data`: you have to specify the data frame with the data you want to plot.

- The `mapping`: the mapping tells ggplot how to map data to parts of the plot like the axes or groupings within the data. These parts are called *aesthetics*, or `aes` for short.

You can specify the data and mapping with the `data` and `mapping` arguments of the `ggplot()` function.

Note that the `mapping` argument is always specified with `aes()`.

In the following bare plot, we are just mapping the total time of sleep (`sleep_total`) to the *x*-axis, and the time of REM (`sleep_rem`) to the *y*-axis.

```{r sleep, exercise=TRUE, exercise.setup = "tidy-load"}
ggplot(
  data = msleep,
  mapping = aes(
    x = sleep_total,
    y = sleep_rem
  )
)
```

### Let's add geometries

Nice, but we are missing the most important part: showing the data!

Data is represented with *geometries*, or `geom`s for short.
`geom`s are added to the base ggplot with functions whose names all start with `geom_`.

For this plot, you want to use `geom_point()`.
This geom simply adds points to the plot based on the data in the `msleep` data frame.

Add the geom to the following code and run it.

Note that the `data` and `mapping` arguments left explicit in the `ggplot()` function, since they are obligatory and they are specified in that order (try running `?ggplot` in the console to see all the arguments of the function).

```{r first-plot, exercise=TRUE, exercise.setup = "tidy-load"}
ggplot(
  msleep,
  aes(sleep_total, sleep_rem)
) +
  ...
```


This type of plot, with two continuous axes and data represented by points, is called a **scatter plot**

## Bar charts

Another common type of plot is the **bar chart**.

Bar charts are useful when you are counting things.
In the following example, we will be counting the number of languages by their endangerment status.

### Do you understand me?

There are thousands of languages in the world, but most of them are loosing speakers, and some are already no longer spoken.
The endangerment status of a language goes from `not endangered` (languages with large populations of speakers) through `threatened`, `shifting` and `nearly extinct`, to `extinct` (languages that have no living speakers left).

The `endangered` data frame contains the endangerment status for 7,845 languages from [Glottolog](https://glottolog.org).
Here's what it looks like.

```{r endangered}
endangered
```

To create a bar chart, add `geom_bar()` to the plot.

You only need one axis, the *x* axis to be precise, because the *y*-axis will always have counts.

```{r endangered-bars, exercise=TRUE, exercise.setup = "tidy-load"}
endangered %>%
  ggplot(aes(status)) +
  geom_bar()
```

Note how the counting is done automatically.
R looks in the `status` column and counts how many times each value in the column occurs in the data frame.

### Forced journey across the Atlantic

European colonisers forced millions of Africans across the Atlantic, to the Americas.
Between 1500 and 1900, more than 12 million people were enslaved and brought to the New World.

The `slavery_tot` data frame contains the number of enslaved people disembarked in the Americas according to digitised historical records, curated by the [Slave Voyages](https://slavevoyages.org) project.

```{r slavery}
slavery_tot
```

Let's create a bar chart with number of enslaved people, separated by the flag of the ship they travelled on.

We can use `geom_bar()` as we did in the previous example, but now there's a catch.

In the `endangered` data frame, `geom_bar()` did the counting of languages for each endangerment status.
But now, the data frame already holds the counts!

Since `slavery_tot` already has counts in the `enslaved` column, we can specify *both* the *x*-axis (`flag`) and the *y*-axis (`enslaved`).
Then we have to tell `geom_bar()` that we have already done the counting: we can do this by specifying `stat = "identity"` (the default is `stat = "count"`).

```{r slavery-bars, exercise=TRUE, exercise.setup = "tidy-load"}
ggplot(
  slavery_tot,
  aes(flag, enslaved)
) +
  ...
```

```{r slavery-bars-hint}
geom_bar(stat = "identity")
```


## Fill and colour

So far, we have used the `aes`thetics `x` for the *x*-axis and `y` for the *y*-axis.

Among the many other `aes`thetics you can specify, there are `fill` and `colour`.

### Fill

The `fill` aesthetic let's you fill bars or areas with different colours depending on the values of a specified column.

Let's recreate the plot on colonial slavery, this time adding some colour. Complete the following code by specifying that `fill` should be based on `country`.

```{r slavery-fill, exercise=TRUE, exercise.setup = "tidy-load"}
slavery_dis %>%
  ggplot(
    aes(flag, count, ...)
  ) +
  geom_bar(stat = "identity")
```

We can improve this plot by reordering the flags based on descending number of disembarked people, with `reorder(flag, desc(count))`.

```{r slavery-fill-ord, exercise=TRUE, exercise.setup = "tidy-load"}
slavery_dis %>%
  ggplot(
    aes(
      reorder(flag, desc(count)),
      count,
      fill = country
    )
  ) +
  geom_bar(stat = "identity") +
  labs(x = "Flag", y = "Number of people (millions)")
```

What if we want to move the colour legend to the bottom of the plot?

Check out the documentation of `theme` by typing `?theme` in the RStudio console and press enter.
Search for the word `position`.

```{r legend-pos, echo=FALSE}
question(
  "Which of the following moves the legend to the bottom of the plot?",
  answer('`legend("bottom")`'),
  answer('`theme(legend.position = "bottom")`', correct = TRUE),
  answer('`theme(legend.bottom)`')
)
```

### Move the legend

Now move the legend to the `"bottom"` of the chart.

```{r slavery-legend, exercise=TRUE, exercise.setup = "tidy-load"}
slavery_dis %>%
  ggplot(
    aes(
      reorder(flag, desc(count)),
      count,
      fill = country
    )
  ) +
  geom_bar(stat = "identity") +
  labs(x = "Flag", y = "Number of people (millions)") +
  ...
```

```{r slavery-legend-hint}
theme(legend.position = "...")
```


### Colour

`colour` works like fill, but it is used with borders or lines of certain geometries like `geom_point()` and `geom_line()`.

In the following plot, we wan to visualise the change in use of land across the globe through time.
There are three uses in data frame: `built_up` for construction land, `cropland` for land use for crops, and `grazing` for land used in animal foraging.
Land `area` is in billion hectares.

Add the aesthetics to the following code and check the resulting plot.
The plot needs `year` and `area` as the axes and `use` for the `colour` aesthetics.

```{r land-use, exercise=TRUE, exercise.setup = "tidy-load"}
land_use %>%
  ggplot(aes(..., ..., ...)) +
  geom_line(size = 2)
```

## Line type

If we wish to use a different type of line depending on the values of a column, we can use the `linetype` aesthetic.

Change the plot on land use so that it uses both `colour` and `linetype` to differentiate the type of land use.

```{r land-use-line, exercise=TRUE, exercise.setup = "tidy-load"}
land_use %>%
  ggplot(aes(..., ..., ...)) +
  geom_line(size = 2)
```

### The end!

Congratulations! You completed this tutorial.
